### 2020-06-08 chapters/background.tex
Reviewed Intro and Background. Added PDIS pdf

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine. For instance, grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark \cite{spark} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times. As more and more data became available in the data stream format, the need for stream processing increased. Stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some computations are impossible to compute accurately on unbounded data sets. Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.

In order to provide a streaming solution, Apache Spark created the Spark Streaming module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not really exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

% streaming as a proper superset of batch
Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work. The Venn diagram \ref{fig:stream-superset} was taken from \cite{batch-is-a-special-case-of-streaming} and illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the content of an unbounded data stream contains that of a bounded data set. That is, from an infinite stream, it is possible to fix a left and right boundary in order to obtain a bounded and finite data set. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from \cite{batch-is-a-special-case-of-streaming}.

@@

%\section{Anatomy of data analytics}
% data stream is composed by, that may exist or not
%window
%aggregation
%group_by

@@

% boundless aggregations no need for windows
Computations over an unbounded data set are not impossible. For example, computing the \textit{average} of an event's field for all incoming events is computationally feasible. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some logical context into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. Is the average of all temperatures of all machines over an infinite period useful for the detection of overheating equipment? In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end. By creating such a conceptual range, we realize that in practice we need a bounded data set. Partitioning an infinite data stream to obtain such a finite data set is made through the use of windows.

The authors of \cite{Wang-Windows-Stream-Processing} claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming aggregations"}. More formally, the authors of \cite{Botan-SECRET} state that \textit{"a window W over a stream S is a finite subset of S"}.

% 2 dimensions of windows
Windows can be categorized by two dimensions: the window size \textit{W} and the sliding step size \textit{S}, both either tuple or time-based. Tuple-based windows' size is specified by the number of items the window can contain and time-based windows specify the time interval worth of incoming data a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is formally defined by \cite{Botan-SECRET} as \textit{"the distance between consecutive windows"}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

@@

For the scope of this Thesis, we will work under a time-based sliding window framework with a unitary step size. Despite focusing on time-based true sliding windows, our approach is just as valid for tuple sized windows and for other step values as well.

@@

Aggregations are present in most programming languages and frameworks. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value for the data set.

@@

While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques. In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs). Considering the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the total sum plus the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed this far minus the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Figure \ref{fig:pseudo-soe} shows pseudocode for SOE \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

@@

To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

>>>>

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine. For instance, grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop\footnote{https://hadoop.apache.org} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark\footnote{https://spark.apache.org} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times.

As more and more data became available in the data stream format, the need for stream processing increased. Stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some aggregations are impossible to compute with 100\% accuracy in an efficient manner on unbounded data sets. An example of such is counting distinct elements that flow through the data stream. An immediate solution would be to create a regular set of items or strings and have its length be the number of distinct elements processed. However, this approach is not very efficient and has a linear memory usage growth regarding the number of distinct elements. Therefore, as the number of distinct elements approaches large orders of magnitude we run out of memory.

In order to provide a streaming solution, Apache Spark created the Spark Streaming\footnote{https://spark.apache.org/docs/latest/streaming-programming-guide.html} module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not really exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work. The Venn diagram \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded data set. That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite data set. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} \cite{batch-is-a-special-case-of-streaming}.

@@

@@

% boundless aggregations no need for windows
Computations over an unbounded data set are not impossible. For example, computing the \textit{average} of an event's field for all incoming events is computationally feasible. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some logical context into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. Is the average of all temperatures of all machines over an infinite period useful for the detection of overheating equipment? In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}. By creating such a conceptual range, we realize that in practice we need a bounded data set. Partitioning an infinite data stream to obtain such a finite data set is made through the use of windows.

Gedik et. al claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming a\cite{Wang-Windows-Stream-Processing}ggregations"} \cite{Wang-Windows-Stream-Processing}. More formally, Botan et. al define that \textit{"a window W over a stream S is a finite subset of S"} \cite{Botan-SECRET}.

% 2 dimensions of windows
Windows can be categorized by two dimensions: the window size \textit{W} and the sliding step size \textit{S}, both either tuple or time-based. Tuple-based windows' size is specified by the number of items the window can contain and time-based windows specify the time interval worth of incoming data a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is formally defined as \textit{"the distance between consecutive windows"} \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

@@

For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size. Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and for other step values as well.

@@

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value.

@@

While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques. In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. Tangwongsan et. al define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. Consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink\footnote{https://flink.apache.org}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Figure \ref{fig:pseudo-soe} shows pseudocode for SOE \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

@@

To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

### 2020-06-08 chapters/introduction.tex
Reviewed Intro and Background. Added PDIS pdf

Feedzai's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained before deployment making assumptions on the statistical distributions of future incoming data. Thus their model performance is linked to the maintenance of these assumptions. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, the performance decay caused by data pattern shifts is also known as concept drift. %UNCOMMENT WHEN BACKGROUND SECTION IS DONE: Such a concept will be further explained in Section \ref{sec:concept-drift}.

@@

The objective of this Thesis is to build a lightweight real-time system that detects these shifts in data patterns in systems that handle high volumes of high velocity, highly skewed and seasonal data.

\section{Motivation} \label{sec:motivation}
For any organization, the quality of their provided services is paramount. Therefore, it is common for companies to have systems monitoring the former, testing their usage rate, availability and performance, just to name a few. The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance.

@@

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.

\section{Hypothesis Validation} \label{sec:validation}
The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item high precision ratio --- \textit{i.e.} at least 80\%
    \item high recall ratio --- \textit{i.e.} at least 80\%
    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
\end{itemize}

Experiments will be made using multiple real data sets from the financial fraud space and compared to existing batch analysis results.

\section{Document structure}
This section merely sums up how each of the following Chapters contributes to this Thesis.

@@

In Chapter \ref{chap:conclusion} we summarize our contributions.

>>>>

Feedzai's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained before deployment making assumptions on the statistical distributions of future incoming data. Thus their model performance is linked to the maintenance of these assumptions. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, the performance decay caused by data pattern shifts is also known as concept drift.

@@

The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relativily to a reference period.

\section{Motivation} \label{sec:motivation}
For any organization, the quality of their provided services is paramount. Therefore, it is common for companies to have systems monitoring the former, testing their usage rate, availability and performance, just to name a few. The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance of the system analyzing said data stream.

@@

\section{Hypothesis} \label{sec:hypothesis}
%The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.

The question this Thesis aims to answer is whether a method that relies on comparing an aggregation of the reference period and the aggregation effectively seen during the stream analysis can be used to detect data pattern shifts in the stream, in the presence of high volumes of high velocity, highly skewed and seasonal data while ensuring a low memory footprint.

\section{Hypothesis Validation} \label{sec:validation}
The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item high recall ratio --- \textit{i.e.} at least 80\%
    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
\end{itemize}

Experiments will be made using multiple synthetic and real data sets from the financial fraud space.

\section{Document structure} 
% TODO at the end rewrite and relink this
This section merely sums up how each of the following Chapters contributes to this Thesis.

@@

%In Chapter \ref{chap:conclusion} we summarize our contributions.
